From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-02 Pi+/paper/finite.tex
Fix text for the Lehmer-Fin equivalence proof

We can also establish an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.

@@

\todo{So now, finally, give an explicit example of a normal form, Lehmer immersion}.

@@

\begin{proposition}~\label{prop:fin-finexcept}
  \leavevmode
  \begin{enumerate}
    \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-1}
    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$.~\label{prop:fin-finexcept-2}
    \item For any $n : \Nat$,
          \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]} \).~\label{prop:fin-finexcept-3}
  \end{enumerate}
\end{proposition}

@@

  The third proposition performs some interesting computations. On the left, we have the type of automorphisms of
  $\Fin[\suc[n]]$. Assume a particular one $\phi : \Fin[\suc[n]] \xrightarrow{\sim} \Fin[\suc[n]]$. To construct an
  element of the pair on the right, we pick the first component to be the image of the element $\zero$ under $\phi$.
  After fixing $k \defeq \phi(\zero)$, we need to compute the bijection in the second component. We remove the two
  elements $\zero$ and $k$ from $\Fin[\suc[n]]$ and construct an equivalence between the resulting types
  $\FinExcept{\suc[n]}[\zero]$ and $\FinExcept{\suc[n]}[k]$ as follows:
  $\FinExcept{\suc[n]}[\zero] \eqv \dsum{k:\Fin[\suc[n]]}{(\phi(\zero) \neq \phi(k))} \eqv \FinExcept{\suc[n]}[k]$.

  To go the other way, given a $k : \Fin[\suc[n]]$ and an equivalence
  $\FinExcept{\suc[n]}[\fzero] \eqv \FinExcept{\suc[n]}[k]$ we compute an equivalence as follows:
  $\Fin[\suc[n]] \eqv \unit \sqcup \FinExcept{\suc[n]}[k] \eqv \unit \sqcup \Fin[n] \eqv \Fin[\suc[n]]$.
\end{proof}

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]          \\
      \eqv & \unit                   \\
      \eqv & \Aut[\Fin[\suc[\zero]]] \\
    \end{array}
    \qquad\qquad
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                              &                                                                 \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                         & \text{by definition}                                            \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                & \text{induction hypothesis}                                     \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                              & \text{$\Sigma$ over a constant family}                          \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \Fin[\suc[m]]}          & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[\fzero] \eqv \FinExcept{\suc[m]}[k]} & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                     & \text{by~\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-3}} \\
    \end{array}
  \end{gather*}
\end{proof}

To see how the proof works combinatorically, let us go through an example. We will start from a following bijection $p: \Aut[\Fin[4]]$:
\begin{align*}
  p(0) \mapsto & 1 \\
  p(1) \mapsto & 3 \\
  p(2) \mapsto & 0 \\
  p(3) \mapsto & 2
\end{align*}

At first, we start from a full domain $\Fin[4]$ and full codomain $\Fin[4]$.
\[
  \begin{bmatrix}
    0 \\
    1 \\
    2 \\
    3
  \end{bmatrix}
  \mapsto
  \begin{bmatrix}
    0 \\
    1 \\
    2 \\
    3
  \end{bmatrix}
\]

We remember that the last element of result tuple is $k - 1 = 2$. The appropriate elements on both sides are removed,
forming, respectively, $\FinExcept{4}[0]$ and $\FinExcept{4}[1]$. These are then
both transformed to $\Fin[3]$ using
\cref{prop:fin-finexcept}.~\ref{prop:fin-finexcept-2}, as shown below.

\[
  \begin{bmatrix}
    0 \\
    1 \\
    2
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    1 \\
    2 \\
    3
  \end{bmatrix}
  \mapsto
  \begin{bmatrix}
    0 \\
    2 \\
    3
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    0 \\
    1 \\
    2
  \end{bmatrix}
\]

Then, the original permutation specified that the element $\el{1}$ has to be
mapped to $\el{3}$. Since we performed renumbering, now it means that element is
mapped to $\el{2}$ - the resulting tuple now looks like $[2, 1]$. We yet again remove these from the domain and codomain,
forming $\FinExcept{0}[0]$ and $\FinExcept{0}[2]$.

\[
  \begin{bmatrix}
    0 \\
    1
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    1 \\
    2
  \end{bmatrix}
  \mapsto
  \begin{bmatrix}
    0 \\
    2
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    0 \\
    1
  \end{bmatrix}
\]

The situation repeats two more times: first, by assinging $\el{0}$ to $\el{1}$,
giving the tuple $[0, 2, 1]$ and transforming the domain and codomain:
\[
  \begin{bmatrix}
    0
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    1
  \end{bmatrix}
  \mapsto
  \begin{bmatrix}
    0
  \end{bmatrix}
  \equiv
  \begin{bmatrix}
    0
  \end{bmatrix}
\]

And finally, by assinging $\el{0}$ to $\el{0}$, forming a tuple $[0, 0, 2, 1]$ and

At this point we see that the renumbering of the domain is done in an orderly
way (always shifting by 1), so it can be ignored. On the other hand, the
ordering of the codomain is what matters. Each time we remove an element, all
larger elemens are in effect decreased by one.

To remind ourselves, when we introduced the combinatorial interpretation of
Lehmer codes, the element $\el{r}$ at postition $k$ was said to count the number
of inversions of the element $\el{r}$. Thus, this is exactly the meachanism in
which the induction step works.

This corresponds exactly to what happens in
the \cref{prop:fin-finexcept-3}. Looking at the operation combinatorially, it is
inserting element $\el{r}$ at the position $k$, meaning that it shifts the
element $k$places.\jk{needs rewriting after the example is here}

Viewing the codomain as a vector of four elements, element $\el{0}$ has zero
inverses, element $\el{1}$ has 1 inverse, element $\el{2}$ has 1 inverse, and
element $3$ has

>>>>

We also established an equivalence between $\im{\normf}$ and $\Sn$, which gives the following.

@@

@@

\begin{proposition}~\label{prop:fin-finexcept}
  \leavevmode
  \begin{enumerate}
    % \item For any $k : \Fin[n]$, $\unit \sqcup \FinExcept{n}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-1}
    \item For any $k : \Fin[\suc[n]]$, $\FinExcept{\suc[n]}[k] \eqv \Fin[n]$. \label{prop:fin-finexcept-2}
    \item For any $n : \Nat$, \( \Aut[\Fin[\suc[n]]] \eqv \dsum{k : \Fin[\suc[n]]}{\FinExcept{\suc[n]}[n] \eqv
            \FinExcept{\suc[n]}[n - k]} \). \label{prop:fin-finexcept-3}
  \end{enumerate}
\end{proposition}

@@

  The third proposition performs some combinatorial tricks. On the left, we have the type of automorphisms of
  $\Fin[\suc[n]]$. Assume a particular $\sigma : \Fin[\suc[n]] \xrightarrow{\sim} \Fin[\suc[n]]$. Pick $k$ to be the
  inversion count of $n$, the largest element in $\Fin[\suc[n]]$. Then, the image of $n$ under $\phi$ has to be $n - k$,
  since all other elements in the set are smaller. Removing those two from the domain and codomain of $\phi$, the rest
  of the elements are fixed by $\sigma$, so we compute the bijection between the rest of the elements.

  For the other direction, if we are given a $k$ and a bijection $\pi$ between $\FinExcept{\suc[n]}[n]$ and
  $\FinExcept{\suc[n]}[n - k]$, we can extend $\pi$ to $\sigma : \Fin[\suc[n]] \eqv \Fin[\suc[n]]$ by inserting the
  element $n$ at the position $n - k$, resulting in the element $n$ having inversion count $k$.
\end{proof}

@@

\begin{proof}
  For $n = \zero$, note that $\Lehmer[\zero]$ is contractible, and so is $\Aut[\Fin[\suc[\zero]]]$. For $n = \suc[m]$,
  we compute a chain of equivalences.
  \begin{gather*}
    \arraycolsep=0.5em\def\arraystretch{1.5}
    \begin{array}{rl}
           & \Lehmer[\zero]          \\
      \eqv & \unit                   \\
      \eqv & \Aut[\Fin[\suc[\zero]]] \\
    \end{array}
    \qquad\qquad
    \begin{array}{rlr}
           & \Lehmer[\suc[m]]                                                                                   &                                                                 \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Lehmer[m]                                                              & \text{by definition}                                            \\
      \eqv & \Fin[\suc[\suc[m]]] \times \Aut[\Fin[\suc[m]]]                                                     & \text{induction hypothesis}                                     \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\Fin[\suc[m]] \eqv \Fin[\suc[m]]}                                   & \text{$\Sigma$ over a constant family}                          \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \Fin[\suc[m]]}                    & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \dsum{k : \Fin[\suc[\suc[m]]]}{\FinExcept{\suc[\suc[m]]}[m] \eqv \FinExcept{\suc[\suc[m]]}[m - k]} & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-2}} \\
      \eqv & \Aut[\Fin[\suc[\suc[m]]]]                                                                          & \text{by~\cref{prop:fin-finexcept}~\cref{prop:fin-finexcept-3}} \\
    \end{array}
  \end{gather*}
\end{proof}
